---
title: "LDC Example"
author: "Michael Schramm"
date: "`r Sys.Date()`"
github-repo: mps9506/twri-bacteria-ldc-example
bibliography: bibliography.bib
biblio-style: "apalike"
link-citations: true
url: "https://mps9506.github.io/twri-bacteria-ldc-example/"
---


```{r include=FALSE}
ragg_png = function(..., res = 192) {
  ragg::agg_png(..., res = res, units = "in")
}
knitr::opts_chunk$set(dev = "ragg_png",
                      fig.ext = "png",
                      fig.retina=2)
```


```{r isetup, message=FALSE, warning=FALSE, cache=FALSE, echo=FALSE}
library(dataRetrieval) ## for USGS daily streamflow data
library(tidyverse)
library(janitor) ## cleans up messy variable names
## uncomment the line below to install
# install.packages(remotes)
# remotes::install_github("mps9506/dartx")
library(dartx)
library(DescTools)
```


# Step 1 {-}

Setup the project in R Studio. Use "File -> New Project", name your project and select the directory to store your project. I always recomend setting up a couple of folder for all project workflows, this can be adjusted based on project needs but this provides a good starting point:

```
Home Folder
  |
  |-RawData
  |-ExportData
  |-Figures
  |-Scripts
  projectname.Rproj
  readme.md
  
```

The idea in this file structure is that we never want to alter the raw data we are manipulating. All data manipulation is done via the scripts and stored in memory and written to a different exported data file. 


# Step 2 {-}

Install and load the required packages. Use `install.packages(dataRetrieval)` as needed to install required packages.

```{r packages}
library(dataRetrieval) ## for USGS daily streamflow data
library(tidyverse)
library(janitor) ## cleans up messy variable names
## uncomment the line below to install
# install.packages(remotes)
# remotes::install_github("mps9506/dartx")
library(dartx)
library(DescTools)


```


# Step 3 {-}

We have two general scenarios we need to consider. The easiest is when there is a monitoring station right next to a stream flow gage. This way we can estimate bacteria loads (count units) using measured streamflow (water volume) and measured bacteria concentration. Concentration times volume equals counts or mass.

The second scenario is when there is no streamflow data where the bacteria data is collected. We need to estimate the streamflow at the water quality site. If there is a stream gage upstream or downstream of our stie of interest, a simple drainage area ratio works well. the `dartx` package provides an empriically measured adjustment factor to the drainage area ratio as applied in Texas.

One caveat, if this is being done for a TMDL (or TSD in preperation for a TMDL), we have to account for wastewater permits, future growth and water rights on streamflow. I am not concerned about that for this example.

Start with the basic example. This example uses Tres Palacios Creek SWQM station 12517. There is a USGS gage, 08162600 located at the site. The bacteria data has been downloaded manually from SWQMIS and is in pipe delmminated format in the RawData folder. The streamflow data will be downloaded from USGS and stored in the RawData folder for future use.

```{r}
# Read in the bacteria data
df_12517 <- read_delim("RawData/SWQM-12517-P31699.txt", 
    "|", escape_double = FALSE, col_types = cols(Segment = col_character(), 
        `Station ID` = col_character(), `Parameter Code` = col_character(), 
        `End Date` = col_date(format = "%m/%d/%Y"), 
        `End Time` = col_skip(), `End Depth` = col_skip(), 
        `Start Date` = col_skip(), `Start Time` = col_skip(), 
        `Start Depth` = col_skip(), `Composite Category` = col_skip(), 
        `Composite Type` = col_skip(), `Submitting Entity` = col_skip(), 
        `Collecting Entity` = col_skip(), 
        `Monitoring Type` = col_skip(), Comments = col_skip()), 
    trim_ws = TRUE)
df_12517 <- clean_names(df_12517)
df_12517
```

This code chunk will download the streamflow data. I usually write this in a sepearte script so I only have to run it one time and write the output file to the raw data folder. The main script will read the data from the raw data folder instead of trying to download the data every single time we run the script.



```{r}
# download mean daily streamflow and write to csv
Q_df <- readNWISdv(siteNumbers = "08162600",
           startDate = "2000-01-01",
           endDate = "2020-12-31",
           parameterCd = "00060",
           statCd = "00003")
Q_df <- renameNWISColumns(Q_df)
write_csv(Q_df,
          "RawData/meandailyQ_08162600.csv")
as_tibble(Q_df)

```

Always do a quick exploratory plot of the data. Nothing stands out on the plots below. We could do some more exploration but I don't think it is needed here.

```{r}
ggplot(df_12517) +
  geom_point(aes(end_date, value)) +
  scale_y_log10()

ggplot(Q_df) +
  geom_line(aes(Date, Flow))
```

## Flow Duration Curve

Now we make a flow duration curve by calculating the probablity that a flow on any given day will be exceeded.

```{r}
Q_df <- Q_df %>%
  select(Date, Flow) %>%
  mutate(FlowExceedance = 1/(length(Flow)/rank(-Flow))) ## this calculates the prob of flow being exceeded

as_tibble(Q_df)
```

We can plot the FDC now as:

```{r}
ggplot(Q_df) +
  geom_line(aes(FlowExceedance,
                Flow)) +
  scale_y_log10() +
  labs(x = "Proportion Time Flow Exceeded",
       y = "Flow (cfs)")
```


## LDC

The next step is to convert the FDC to an LDC with bacteria data. The steps are to multiply the flow by the instream bacteria standard. This tells us the allowable bacteria loading in the stream under different flow exceedance conditions.

```{r}
Q_df %>%
  ## MPN/100mL * cubic feet/sec * mL/cubic feet * sec/day = mpn/day
  mutate(LDC = (126/100) * Flow * 28316.8 * 86400) -> Q_df

ggplot(Q_df) +
  geom_line(aes(FlowExceedance, LDC)) +
  scale_y_log10() +
  labs(x = "Proportion Time Load Exceeded",
       ## sorry, stupid way to get italics into labels
       y = expression(paste(italic("E. coli"), " (MPN/day)")))
```

Now we want to overlay the measured bacteria data as loads on the LDC. So we pair the bacteria data by date to flow, then convert the concentration to a daily load and add it to the plot.

```{r}
Q_df %>%
  left_join(df_12517 %>%
              select(end_date, value),
            by = c("Date" = "end_date")) %>%
  mutate(MeasuredLoad = (value/100) * Flow * 28316.8 * 86400) -> Q_df

as_tibble(Q_df)
```


```{r}
ggplot(Q_df) +
  geom_line(aes(FlowExceedance, LDC,
                linetype = "Allowable Load at Geomean Criterion (126 MPN/100 mL)")) +
  geom_point(aes(FlowExceedance, MeasuredLoad,
                 shape = "Measurement Value (MPN/day)",
                 color = "Measurement Value (MPN/day)")) +
  scale_y_log10() +
  labs(x = "Proportion Time Load Exceeded",
       y = expression(paste(italic("E. coli"), " (MPN/day)"))) +
  scale_shape_manual(name = "values", values = c(21)) +
  scale_color_manual(name = "values", values = c("dodgerblue4")) +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        legend.title = element_blank(),
        plot.margin = margin(5, 10, 5, 5))
```



Now we need to split the flow exceedance into different categories and summarize the loads by category. Here we make a summary table that we can add to our ggplot.

```{r}
load_summary <- Q_df %>%
    mutate(Flow_Condition = case_when(
    FlowExceedance >= 0 & FlowExceedance < 0.1 ~ "Highest Flows",
    FlowExceedance >= 0.1 & FlowExceedance < 0.4 ~ "Moist Conditions",
    FlowExceedance >= 0.4 & FlowExceedance < 0.6 ~ "Mid-Range Flows",
    FlowExceedance >= 0.6 & FlowExceedance < 0.9 ~ "Dry Conditions",
    FlowExceedance >= 0.9 & FlowExceedance <= 1 ~ "Lowest Flows"
  )) %>%
  mutate(p = case_when(
    FlowExceedance > 0 & FlowExceedance <= 0.1 ~ 0.05,
    FlowExceedance > 0.1 & FlowExceedance <= 0.4 ~ 0.25,
    FlowExceedance > 0.4 & FlowExceedance <= 0.6 ~ 0.5,
    FlowExceedance > 0.6 & FlowExceedance <= 0.9 ~ 0.75,
    FlowExceedance > 0.9 & FlowExceedance <= 1 ~ 0.95
  )) %>%
  group_by(Flow_Condition, p) %>%
    summarise(quantileflow = round(quantile(Flow, .5, type = 5, names = FALSE), 3),
            geomean_sample = DescTools::Gmean(value, na.rm = TRUE),
            calcload = quantileflow * geomean_sample/100 * 28316.8 * 86400)
load_summary
```



## Final plot

Final plot:

```{r}
ggplot(Q_df) +
  geom_vline(xintercept = c(.10, .40, .60, .90), color = "#cccccc") +
  geom_line(aes(FlowExceedance, LDC,
                linetype = "Allowable Load at Geomean Criterion (126 MPN/100 mL)")) +
  geom_point(aes(FlowExceedance, MeasuredLoad,
                 shape = "Measurement Value (MPN/day)",
                 color = "Measurement Value (MPN/day)")) +
  geom_point(data = load_summary, aes(p, calcload, 
                                      shape = "Exisiting Geomean Load (MPN/day)",
                                      color = "Exisiting Geomean Load (MPN/day)")) +
  annotation_logticks(sides = "l", color = "#cccccc") +
  annotate("text", x = .05, y = max(Q_df$MeasuredLoad, na.rm = TRUE) + (0.5 * max(Q_df$MeasuredLoad, na.rm = TRUE)), label = "High\nflows", hjust = 0.5, size = 3, family = "Arial Condensed", lineheight = 1) +
  annotate("text", x = .25, y = max(Q_df$MeasuredLoad, na.rm = TRUE) + (0.5 * max(Q_df$MeasuredLoad, na.rm = TRUE)), label = "Moist\nconditions", hjust = 0.5, size = 3, family = "Arial Condensed", lineheight = 1) +
  annotate("text", x = .50, y = max(Q_df$MeasuredLoad, na.rm = TRUE) + (0.5 * max(Q_df$MeasuredLoad, na.rm = TRUE)), label = "Mid-range\nflows", hjust = 0.5, size = 3, family = "Arial Condensed", lineheight = 1) +
  annotate("text", x = .75, y = max(Q_df$MeasuredLoad, na.rm = TRUE) + (0.5 * max(Q_df$MeasuredLoad, na.rm = TRUE)), label = "Dry\nconditions", hjust = 0.5, size = 3, family = "Arial Condensed", lineheight = 1) +
  annotate("text", x = .95, y = max(Q_df$MeasuredLoad, na.rm = TRUE) + (0.5 * max(Q_df$MeasuredLoad, na.rm = TRUE)), label = "Low\nflows", hjust = 0.5, size = 3, family = "Arial Condensed", lineheight = 1) +
  scale_y_log10() +
  labs(x = "Proportion Time Load Exceeded",
       y = expression(paste(italic("E. coli"), " (MPN/day)"))) +
  scale_shape_manual(name = "values", values = c(12, 21)) +
  scale_color_manual(name = "values", values = c("red", "dodgerblue4")) +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.direction = "vertical",
        legend.title = element_blank(),
        plot.margin = margin(5, 10, 5, 5))
```

# References {-}
